---
title: "Getting started with soilKey"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Getting started with soilKey}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment  = "#>"
)
```

`soilKey` provides automated soil profile classification under WRB 2022 (4th edition) and SiBCS (5th edition). The taxonomic key itself is implemented as deterministic R code driven by versioned YAML rules; vision-language extraction, spatial priors, and OSSL-based attribute prediction sit alongside it as modular layers, never inside it.

This vignette walks through the v0.2 deliverable: building a `PedonRecord`, calling the eleven implemented WRB 2022 horizon and property diagnostics directly, classifying canonical fixtures end-to-end through the wired key, and watching the evidence grade respond to provenance.

```{r load}
library(soilKey)
```

# 1. Building a PedonRecord from scratch

`PedonRecord` is the central data carrier. It bundles site metadata, the horizons table (with a fixed canonical schema -- see `horizon_column_spec()` for the full list of columns), and optional spectra, images, documents, and a per-attribute provenance log.

```{r build-pedon}
my_pedon <- PedonRecord$new(
  site = list(
    id              = "example-001",
    lat             = -22.5,
    lon             = -43.7,
    country         = "BR",
    parent_material = "gneiss"
  ),
  horizons = data.frame(
    top_cm      = c(0,  15, 65, 130),
    bottom_cm   = c(15, 65, 130, 200),
    designation = c("A", "Bw1", "Bw2", "C"),
    clay_pct    = c(50, 60,  65,  60),
    silt_pct    = c(15, 10,  8,   8),
    sand_pct    = c(35, 30,  27,  32),
    cec_cmol    = c(8,  5,   4.5, 4),
    bs_pct      = c(20, 12,  10,  11),
    ph_h2o      = c(4.8, 4.9, 5.0, 5.1),
    oc_pct      = c(2.0, 0.4, 0.2, 0.1)
  )
)

my_pedon$validate()
```

The validator catches inverted depths, texture sums far from 100, implausible pH, sum of bases above CEC, Munsell out-of-range, and a handful of other soil-physical sanity checks.

# 2. Canonical fixtures

`soilKey` ships sixteen canonical fixtures designed so that exactly one of the eleven v0.2 diagnostics passes on each. Each profile also classifies cleanly through the wired WRB key.

```{r fixture-list}
fixtures <- list(
  Ferralsol  = make_ferralsol_canonical(),
  Luvisol    = make_luvisol_canonical(),
  Acrisol    = make_acrisol_canonical(),
  Lixisol    = make_lixisol_canonical(),
  Alisol     = make_alisol_canonical(),
  Chernozem  = make_chernozem_canonical(),
  Kastanozem = make_kastanozem_canonical(),
  Phaeozem   = make_phaeozem_canonical(),
  Calcisol   = make_calcisol_canonical(),
  Gypsisol   = make_gypsisol_canonical(),
  Solonchak  = make_solonchak_canonical(),
  Cambisol   = make_cambisol_canonical(),
  Plinthosol = make_plinthosol_canonical(),
  Podzol     = make_podzol_canonical(),
  Gleysol    = make_gleysol_canonical(),
  Vertisol   = make_vertisol_canonical()
)

ferralsol <- fixtures$Ferralsol
ferralsol
```

# 3. Calling the diagnostics directly

Every diagnostic returns a `DiagnosticResult` carrying the per-sub-test evidence, missing-attribute report, layer indices that satisfied, and the WRB literature reference.

```{r ferralic}
ferralic(ferralsol)
```

# 4. Diagnostic matrix across the canonical fixtures

```{r matrix, results='asis'}
diagnostics <- c("argic", "ferralic", "mollic", "calcic", "gypsic", "salic",
                  "cambic", "plinthic", "spodic",
                  "gleyic_properties", "vertic_properties")

mat <- vapply(fixtures, function(p) {
  vapply(diagnostics, function(d) {
    fn <- get(d, envir = asNamespace("soilKey"))
    isTRUE(fn(p)$passed)
  }, logical(1))
}, logical(length(diagnostics)))

knitr::kable(t(mat))
```

Every fixture activates exactly one diagnostic (or, for the argic-derived RSGs Acrisol / Lixisol / Alisol / Luvisol, just the shared `argic`).

# 5. RSG-derived diagnostics: argic and mollic families

The argic horizon is shared by four RSGs -- Acrisols, Lixisols, Alisols, Luvisols -- which differ by clay activity (CEC per kg clay) and chemistry (BS or Al saturation). soilKey provides one diagnostic per RSG that runs `argic()` internally, then applies the activity and chemistry tests on the argic layer:

```{r argic-derived}
acrisol(make_acrisol_canonical())$passed
lixisol(make_lixisol_canonical())$passed
alisol (make_alisol_canonical())$passed
luvisol(make_luvisol_canonical())$passed
```

Same pattern for the mollic-derived family (Chernozems / Kastanozems / Phaeozems):

```{r mollic-derived}
chernozem (make_chernozem_canonical())$passed
kastanozem(make_kastanozem_canonical())$passed
phaeozem  (make_phaeozem_canonical())$passed
```

# 6. End-to-end WRB classification

`classify_wrb2022()` consumes a `PedonRecord` and runs it through the YAML key (`inst/rules/wrb2022/key.yaml`). v0.2 wires 16 of 32 RSGs end-to-end; the other 16 are stubbed with `not_implemented_v01:` markers and return NA in the trace.

```{r classify-fr}
classify_wrb2022(ferralsol)
```

```{r classify-all}
classifications <- vapply(fixtures, function(p) {
  classify_wrb2022(p, on_missing = "silent")$rsg_or_order
}, character(1))
data.frame(fixture = names(classifications), assigned_rsg = classifications)
```

Each canonical fixture maps to its intended RSG. The trace shows which RSGs were tested, in canonical key order, before the assigned one.

# 7. Provenance and evidence grade

`PedonRecord$add_measurement()` records a value's provenance in a structured log. The final `ClassificationResult$evidence_grade` summarises that log on an A--D scale: A means every recorded value was laboratory-measured, D means the result rests on attributes extracted by VLM or assumed by the user.

```{r provenance}
ferralsol_v <- make_ferralsol_canonical()

# Mark the Bw1 clay value as predicted from spectroscopy
ferralsol_v$add_measurement(
  horizon_idx = 4,
  attribute   = "clay_pct",
  value       = 60,
  source      = "predicted_spectra",
  confidence  = 0.85,
  overwrite   = TRUE
)

classify_wrb2022(ferralsol_v)$evidence_grade
```

```{r provenance-vlm}
ferralsol_w <- make_ferralsol_canonical()
ferralsol_w$add_measurement(1, "clay_pct", 50, "extracted_vlm",
                              confidence = 0.7, overwrite = TRUE)
classify_wrb2022(ferralsol_w)$evidence_grade
```

# 8. Interoperability with `aqp`

`PedonRecord$to_aqp()` returns an `aqp::SoilProfileCollection`, allowing soilKey to plug into aqp's plotting and aggregation tooling without owning that infrastructure:

```{r to-aqp, eval = requireNamespace("aqp", quietly = TRUE)}
spc <- ferralsol$to_aqp()
class(spc)
aqp::profile_id(spc)
```

# 9. v0.2 scope and the v0.3+ roadmap

| Version | Scope                                                                                                                |
|---------|----------------------------------------------------------------------------------------------------------------------|
| v0.1    | Core classes; argic, ferralic, mollic; Ferralsols path                                                               |
| v0.2    | +calcic, gypsic, salic, cambic, plinthic, spodic, gleyic, vertic; +AC/LX/AL/LV/CH/KS/PH RSG diagnostics; 16/32 wired |
| v0.3    | Remaining 16 RSGs (Histosols, Anthrosols, Technosols, Cryosols, Leptosols, Solonetz, Andosols, Nitisols, Planosols, Stagnosols, Umbrisols, Durisols, Retisols, Arenosols, Fluvisols); WRB qualifier resolution |
| v0.4    | Module 4 -- OSSL spectroscopy bridge (MBL, PLSR-local, pretrained)                                                   |
| v0.5    | Module 3 -- SoilGrids spatial prior + Brazilian Embrapa raster                                                       |
| v0.6    | Module 2 -- Multimodal extraction (PDF/photo -> structured horizons via `ellmer`)                                    |
| v0.7    | Module 6 -- Parallel SiBCS key (13 ordens), Embrapa partnership for validation                                        |
| v0.8    | Module 5 -- USDA Soil Taxonomy parallel key (12 orders)                                                              |
| v0.9    | All 202 WRB qualifiers; vignettes 05-09; WoSIS benchmark                                                             |
| v1.0    | CRAN submission and methodological paper                                                                             |

See `ARCHITECTURE.md` (in the package root) for the full design rationale.
