---
title: "Vision-language extraction of pedon data (Module 2)"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Vision-language extraction of pedon data (Module 2)}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment  = "#>"
)
library(soilKey)
```

Module 2 (`vlm-*`) lets `soilKey` build a `PedonRecord` from a field-description PDF, a profile photo, or a fieldsheet image -- using a vision-language model (VLM) for the extraction and JSON-Schema validation as a hard gate. **The taxonomic key itself is never delegated to the LLM**: the VLM is restricted to extraction, and every extracted value is recorded in the provenance log so the evidence grade reflects the lower confidence of VLM-sourced data.

This vignette walks the extraction loop end to end, using a `MockVLMProvider` so the example runs offline and without API keys. Swapping the mock for a real `ellmer` chat (Anthropic, OpenAI, etc.) is a one-line change.

# 1. The mock provider

`MockVLMProvider` exposes the same `$chat()` method as an `ellmer` chat object, but pops responses from a pre-loaded queue. Because all of `soilKey`'s VLM logic talks to the provider through `$chat()`, swapping it for a real chat (or for any custom backend) is transparent.

```{r mock}
mock <- MockVLMProvider$new(responses = list())
class(mock)
```

# 2. Schemas and prompts

VLM responses are constrained by a JSON Schema (Draft-07). The schema for horizon extraction is shipped at `inst/schemas/horizon.json`; the rendered prompt template lives at `inst/prompts/extract_horizons.md`.

```{r schema-peek, eval = requireNamespace("jsonlite", quietly = TRUE)}
sch <- jsonlite::fromJSON(soilKey:::load_schema("horizon"),
                            simplifyVector = FALSE)
length(sch$properties$horizons$items$properties)
head(names(sch$properties$horizons$items$properties), 12)
```

The schema enforces strict types -- `top_cm` must be numeric, `designation` must be a string, Munsell colours come as `{hue, value, chroma, confidence, source_quote}` triples, etc. Any response that fails validation triggers the retry loop with the validation error included in the next prompt.

# 3. Walking the extraction loop with the mock

Suppose the model returns the following horizon-level JSON for a synthetic Latossolo description:

```{r mock-response}
horizon_json <- '{
  "horizons": [
    {
      "top_cm": 0,
      "bottom_cm": 15,
      "designation": "A",
      "munsell_moist": {"hue": "2.5YR", "value": 3, "chroma": 4,
                          "confidence": 0.85, "source_quote": "vermelho-escuro"},
      "clay_pct": {"value": 50, "confidence": 0.9, "source_quote": "muito argilosa (50%)"},
      "oc_pct"  : {"value": 2.0, "confidence": 0.85, "source_quote": "C org. 2.0%"}
    },
    {
      "top_cm": 15,
      "bottom_cm": 65,
      "designation": "Bw1",
      "munsell_moist": {"hue": "2.5YR", "value": 3, "chroma": 6,
                          "confidence": 0.85, "source_quote": "vermelho"},
      "clay_pct": {"value": 60, "confidence": 0.9, "source_quote": "muito argilosa"},
      "oc_pct"  : {"value": 1.2, "confidence": 0.85, "source_quote": "C org. 1.2%"}
    }
  ]
}'
```

Wire the mock provider to return this JSON, send a prompt through `validate_or_retry`, and watch the loop:

```{r validate-or-retry, eval = requireNamespace("jsonlite", quietly = TRUE) && requireNamespace("jsonvalidate", quietly = TRUE)}
mock <- MockVLMProvider$new(responses = list(horizon_json))

res <- soilKey:::validate_or_retry(
  provider    = mock,
  prompt      = "extract horizons from <fake document>",
  schema      = "horizon",
  max_retries = 0L
)

str(res, max.level = 2)
```

The returned `res$data` is the parsed (validated) R object. `res$attempts == 1` because the canned response passed validation on the first try. The `validation_error_at` argument of `MockVLMProvider$new()` forces an attempt to return malformed JSON, exercising the retry path -- see `tests/testthat/test-vlm-extract.R` for a worked retry-loop test.

# 4. Merging extraction into a `PedonRecord`

`apply_horizons_extraction()` consumes the parsed JSON and merges it into a `PedonRecord`, recording each value in the provenance log with `source = "extracted_vlm"`.

```{r apply, eval = requireNamespace("jsonlite", quietly = TRUE) && requireNamespace("jsonvalidate", quietly = TRUE)}
pr <- PedonRecord$new(
  site = list(id = "VLM-demo", lat = -22.5, lon = -43.7,
                country = "BR", parent_material = "gneiss"),
  horizons = data.table::data.table(top_cm = numeric(0), bottom_cm = numeric(0))
)

added <- soilKey:::apply_horizons_extraction(pr, res$data, overwrite = TRUE)
cat("Provenance entries added:", added, "\n")

# Inspect what landed in the horizons table.
pr$horizons[, .(top_cm, bottom_cm, designation,
                munsell_hue_moist, munsell_value_moist, munsell_chroma_moist,
                clay_pct, oc_pct)]
```

# 5. Provenance and evidence grade

Every value that came from the VLM carries `source = "extracted_vlm"` in the provenance log. The classification's evidence grade reacts:

```{r grade, eval = requireNamespace("jsonlite", quietly = TRUE) && requireNamespace("jsonvalidate", quietly = TRUE)}
prov <- pr$provenance
head(prov[, .(horizon_idx, attribute, source, confidence)])
```

Even the simplest call to `classify_wrb2022()` will reflect the lower-confidence VLM source via a lower evidence grade than the same profile assembled from lab data. (See vignette 02 for the grade ladder.)

# 6. Production: swap the mock for an `ellmer` chat

In production, replace `MockVLMProvider` with a chat object from `ellmer`. The rest of the pipeline is unchanged:

```{r ellmer, eval = FALSE}
# install.packages("ellmer")
chat <- ellmer::chat_anthropic(
  model  = "claude-sonnet-4-6",
  system = "You extract structured pedon data from Brazilian field reports."
)

extract_horizons_from_pdf(
  pedon       = pr,
  pdf_path    = "field-reports/perfil-LV-001.pdf",
  provider    = chat,
  max_retries = 3L
)
```

Photos go through `extract_munsell_from_photo()`; site metadata through `extract_site_from_fieldsheet()`. All three follow the same validate-or-retry contract so they accept any provider that exposes `$chat(prompt, ...)`.

The next vignette (`05-spatial-spectra-pipeline`) shows how the SoilGrids spatial prior and OSSL Vis-NIR predictions can fill remaining gaps in the pedon before classification.
