---
title: "Benchmarking soilKey against WoSIS"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Benchmarking soilKey against WoSIS}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment  = "#>"
)
library(soilKey)
```

This vignette describes the benchmark protocol that measures `soilKey`'s WRB 2022 classification agreement against an external reference set, with **WoSIS** (the World Soil Information Service, ISRIC) as the canonical target. It also runs a fully reproducible mini-benchmark on the package's own canonical fixtures so the protocol can be exercised offline.

The full WoSIS run (see §5) is the basis of the agreement statistics reported in the methodological paper accompanying `soilKey` v1.0.

# 1. Protocol overview

Each benchmark profile is a triple $(P, T, C)$:

* $P$ -- a `PedonRecord` assembled from the source dataset's horizons + chemistry;
* $T$ -- the **target** RSG name from the source's own classification (e.g. WoSIS' `cwrb_reference_soil_group`);
* $C$ -- the **candidate** classification produced by `classify_wrb2022(P)`.

The benchmark reports four numbers per source:

* **Top-1 agreement**     -- fraction of profiles where $C\$rsg = T$.
* **Top-3 agreement**     -- fraction where $T$ is among the three highest-likelihood RSGs in the trace (key trace + spatial prior, when available).
* **Coverage**            -- fraction of profiles where $C$ is a definite RSG (not "Regosols" catch-all).
* **Indeterminate-rate**  -- fraction of profiles where $C\$rsg$ is NA owing to missing diagnostic attributes.

Confusion matrices are reported alongside the headline numbers: which RSGs `soilKey` confuses with which (e.g. Acrisol vs Lixisol differ only by base saturation, so disagreement on this border is informative).

# 2. Mini-benchmark on the canonical fixtures

The package's 31 canonical fixtures provide a self-contained benchmark: each fixture is *constructed* to map to a specific RSG, so the expected target is known. The mini-benchmark verifies that `classify_wrb2022()` agrees with the construction intent on every fixture -- i.e. top-1 agreement is 100% on this set.

```{r mini-benchmark}
expected <- c(
  HS = "Histosols", AT = "Anthrosols", TC = "Technosols", CR = "Cryosols",
  LP = "Leptosols", SN = "Solonetz",   VR = "Vertisols", SC = "Solonchaks",
  GL = "Gleysols",  AN = "Andosols",   PZ = "Podzols",   PT = "Plinthosols",
  PL = "Planosols", ST = "Stagnosols", NT = "Nitisols",  FR = "Ferralsols",
  CH = "Chernozems", KS = "Kastanozems", PH = "Phaeozems", UM = "Umbrisols",
  DU = "Durisols",  GY = "Gypsisols", CL = "Calcisols", RT = "Retisols",
  AC = "Acrisols",  LX = "Lixisols",   AL = "Alisols",   LV = "Luvisols",
  CM = "Cambisols", AR = "Arenosols",  FL = "Fluvisols"
)

fixfns <- list(
  HS = make_histosol_canonical,  AT = make_anthrosol_canonical,
  TC = make_technosol_canonical, CR = make_cryosol_canonical,
  LP = make_leptosol_canonical,  SN = make_solonetz_canonical,
  VR = make_vertisol_canonical,  SC = make_solonchak_canonical,
  GL = make_gleysol_canonical,   AN = make_andosol_canonical,
  PZ = make_podzol_canonical,    PT = make_plinthosol_canonical,
  PL = make_planosol_canonical,  ST = make_stagnosol_canonical,
  NT = make_nitisol_canonical,   FR = make_ferralsol_canonical,
  CH = make_chernozem_canonical, KS = make_kastanozem_canonical,
  PH = make_phaeozem_canonical,  UM = make_umbrisol_canonical,
  DU = make_durisol_canonical,   GY = make_gypsisol_canonical,
  CL = make_calcisol_canonical,  RT = make_retisol_canonical,
  AC = make_acrisol_canonical,   LX = make_lixisol_canonical,
  AL = make_alisol_canonical,    LV = make_luvisol_canonical,
  CM = make_cambisol_canonical,  AR = make_arenosol_canonical,
  FL = make_fluvisol_canonical
)

bench <- do.call(rbind, lapply(names(expected), function(code) {
  fx <- fixfns[[code]]()
  res <- classify_wrb2022(fx, on_missing = "silent")
  data.frame(
    fixture  = code,
    target   = expected[[code]],
    assigned = res$rsg_or_order,
    name     = res$name,
    grade    = res$evidence_grade,
    match    = res$rsg_or_order == expected[[code]]
  )
}))

knitr::kable(bench[, c("fixture", "target", "assigned", "match", "grade")])
```

Headline numbers:

```{r mini-stats}
data.frame(
  metric = c("n_profiles", "top1_agreement",
             "indeterminate_rate", "evidence_grade_A",
             "evidence_grade_B"),
  value  = c(nrow(bench),
             mean(bench$match),
             mean(is.na(bench$assigned)),
             mean(bench$grade == "A"),
             mean(bench$grade == "B"))
)
```

Confusion matrix on the mini set:

```{r mini-confusion}
table(target = bench$target, assigned = bench$assigned)
```

# 3. Confusion at the qualifier level

Beyond the RSG, `classify_wrb2022()` returns the principal- and supplementary-qualifier lists. The benchmark can be deepened to track agreement on the **qualifier prefix** of the soil name (the strongest principal that fired). For the canonical fixtures:

```{r mini-qualifier}
qualifier_prefix <- vapply(names(expected), function(code) {
  fx <- fixfns[[code]]()
  res <- classify_wrb2022(fx, on_missing = "silent")
  if (length(res$qualifiers$principal) == 0) NA_character_
  else res$qualifiers$principal[1]
}, character(1))

knitr::kable(
  data.frame(fixture          = names(qualifier_prefix),
             principal_prefix = qualifier_prefix),
  caption = "Most-specific principal qualifier per canonical fixture."
)
```

# 4. Benchmark protocol against WoSIS (run-once, paper-grade)

The mini-benchmark above proves the package's internal consistency. The paper-grade benchmark runs the same flow against ISRIC WoSIS, a global archive of ~118 000 georeferenced pedon descriptions with WRB 2014 / 2022 classifications.

The protocol is intentionally separated from the package itself (no live download in vignettes) so it can be re-run at a controlled date. The driver script is shipped at `inst/benchmarks/run_wosis_benchmark.R`. Skeleton:

```{r wosis-protocol, eval = FALSE}
library(soilKey)

# 1. Pull a snapshot of WoSIS profiles via the WoSIS API.
profiles <- read_wosis_profiles(
  url     = "https://wosis.isric.org/api/v3/profiles?format=json",
  page_size = 500L
)

# 2. Build a PedonRecord per profile.
pedons <- lapply(profiles, build_pedon_from_wosis)

# 3. Classify each through the v0.9.3 key.
classifications <- lapply(pedons, classify_wrb2022, on_missing = "silent")

# 4. Compare against the WoSIS-recorded RSG.
bench <- mapply(function(c, p) {
  data.frame(
    profile_id  = p$site$id,
    target_rsg  = p$site$wosis_rsg,
    assigned    = c$rsg_or_order,
    grade       = c$evidence_grade,
    match       = c$rsg_or_order == p$site$wosis_rsg
  )
}, classifications, pedons, SIMPLIFY = FALSE)
bench <- do.call(rbind, bench)

# 5. Headline numbers.
list(
  n              = nrow(bench),
  top1           = mean(bench$match, na.rm = TRUE),
  indeterminate  = mean(is.na(bench$assigned)),
  pct_grade_A    = mean(bench$grade == "A"),
  by_rsg         = table(bench$target_rsg, bench$assigned)
)
```

The same flow extends naturally to:

* **WoSIS subsets by region** -- e.g. South America only, or Brazilian profiles only (most relevant for the SiBCS cross-system check).
* **WoSIS profiles missing one or more attributes** -- exercise the gap-filling pipeline (vignette 05) and report the agreement uplift from `predicted_spectra`.
* **WoSIS profiles labelled with v2014 vs v2022 RSGs** -- the benchmark separates them so `soilKey`'s v2022-only key is not penalised when WoSIS used a deprecated RSG name.

# 5. Reporting standards

Each benchmark run produces a versioned report at `inst/benchmarks/reports/wosis_<DATE>.md` containing:

* Snapshot date and WoSIS subset used.
* Top-1 / Top-3 agreement, with 95% CI from a non-parametric bootstrap.
* Confusion matrix (RSG-level) and qualifier-prefix agreement.
* Per-RSG agreement and the most common confused pair for each.
* Evidence-grade distribution.
* Notable disagreements with diagnostic-level traces, for QA.

The same script is used for the periodic regression run that ships with each minor `soilKey` release.

# Summary

`soilKey`'s benchmarking layer is intentionally minimal:

1. The mini-benchmark on canonical fixtures is fully reproducible inside this vignette and acts as a continuous-integration check that the key + qualifier system holds 100% top-1 agreement on profiles constructed to match each RSG.
2. The full WoSIS benchmark is run-once-per-release, off-line, with the driver script and reports versioned under `inst/benchmarks/`.
3. Both layers report by the same metrics, so comparison across releases is direct.

This closes the methodological loop opened in the architecture document: the deterministic key, the side modules (VLM, spatial, OSSL), and the benchmark are integrated without any of them being able to overrule the others.
