---
title: "Phase 1 — VLM extraction benchmark"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Phase 1 — VLM extraction benchmark}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>", eval = FALSE)
```

## Why measure first?

Before deciding to **fine-tune** Gemma 4 with a LoRA adapter (Phase 3)
or **inject few-shot demonstrations** into every call (Phase 2), we
need to know how good the **vanilla local Gemma 4 + soilKey persona**
already is. Phase 1 is exactly that measurement.

`benchmark_vlm_extraction()` runs each VLM provider against a small
fixture set covering three soilKey extraction tasks and reports a per-
task quality score, plus an `ok / total` reliability ratio.

| Task | Input | Metric (lower-better unless noted) |
|------|-------|-------------------------------------|
| `horizons` | Markdown / text profile description | precision + recall + per-attribute match (higher-better) |
| `site`     | Field-sheet text | IoU + value-accuracy + recall (higher-better) |
| `munsell`  | Profile photo (with Munsell card) | mean CIE ΔE 2000 (lower-better) |

## Quick start

```{r}
library(soilKey)

# Local Gemma 4 e2b (1.5 GB) -- assumes setup_local_vlm("light") was run.
bench <- benchmark_vlm_extraction(
  providers = list(
    gemma_e2b = list(name = "ollama", model = "gemma4:e2b")
  ),
  tasks         = c("horizons", "site"),  # munsell needs photo fixtures
  max_per_task  = NULL,                    # use every bundled fixture
  verbose       = TRUE
)

bench$summary
```

Compare two locals + a cloud reference:

```{r}
bench <- benchmark_vlm_extraction(
  providers = list(
    gemma_e2b = list(name = "ollama", model = "gemma4:e2b"),
    gemma_e4b = list(name = "ollama", model = "gemma4:e4b"),
    claude    = list(name = "anthropic")
  ),
  tasks = c("horizons", "site")
)
bench$summary
```

## Bundled fixtures (v0.9.66)

```
inst/fixtures/vlm_extraction/
+-- horizons/
|   +-- perfil_RJ_argissolo.txt
|   +-- perfil_RJ_argissolo.golden.json
|   +-- perfil_MG_latossolo.txt
|   `-- perfil_MG_latossolo.golden.json
+-- site/
|   +-- ficha_RJ_001.txt
|   +-- ficha_RJ_001.golden.json
|   +-- ficha_MG_002.txt
|   `-- ficha_MG_002.golden.json
`-- munsell/
    `-- README.md          # see below for the photo-fixture format
```

Each task directory holds **paired** `(input, <id>.golden.json)`
files. The bundled fixtures are deliberately small and synthetic:
short, clean PT-BR profile descriptions with a known answer. Real
benchmarking should add real BDsolos / FEBR cases (instructions
below).

## Adding fixtures from real BDsolos pedons

`make_synthetic_horizons_fixture()` renders any `PedonRecord` back
into a Markdown description and emits the structured horizons as the
golden answer. The combination forms a valid `horizons` fixture:

```{r}
pedons <- load_bdsolos_csv("RJ.csv")
ped    <- pedons[[1L]]

make_synthetic_horizons_fixture(
  ped,
  fixture_id = "bdsolos_RJ_perfil_001",
  out_dir    = system.file("fixtures", "vlm_extraction", "horizons",
                            package = "soilKey")
)

# Re-run the benchmark with the new fixture included:
benchmark_vlm_extraction(
  providers    = list(gemma_e2b = list(name = "ollama", model = "gemma4:e2b")),
  tasks        = "horizons"
)
```

For Munsell-from-photo, add real images plus their hand-curated golden
JSON to `inst/fixtures/vlm_extraction/munsell/` -- see the README in
that directory for the schema and ΔE interpretation.

## Baseline (v0.9.66, gemma4 8B, MacBook M1)

Smoke run on the four bundled text fixtures:

| Provider | Task | Precision / IoU | Recall / value-acc | Attr-match |
|----------|------|-----------------|--------------------|------------|
| `gemma4` (8B, local) | horizons (Latossolo MG) | 1.00 | 1.00 | 1.00 |
| `gemma4` (8B, local) | horizons (Argissolo RJ) | 1.00 | 1.00 | 1.00 |
| `gemma4` (8B, local) | site (Ficha MG)         | 0.79 | 1.00 | 0.79 |
| `gemma4` (8B, local) | site (Ficha RJ)         | 0.87 | 0.92 | 0.87 |

Reads:

- **Horizons extraction is solved** for clean text profiles.
  Precision = recall = 1.0; every numeric attribute (clay %, pH,
  C-org) matched within 10 % tolerance.
- **Site extraction is good but incomplete**: the model gets ~83 %
  of fields and ~96 % of the matched fields agree on value -- but
  fields mentioned only obliquely (`country: BR` inferred from
  state) are missed by the smaller model.

These numbers establish the **vanilla baseline**. Phase 2 and Phase 3
should target the gaps:

- **Phase 2 (few-shot)**: inject 2-3 demonstration pairs per call to
  recover the missing site fields. ~2 days of human work; no GPU.
- **Phase 3 (LoRA)**: train a Parameter-Efficient adapter on (input,
  golden) pairs from BDsolos + FEBR. Needs ~1 000 labelled pairs and
  ~6 h on a single H100. Only justified if Phase 2 plateaus.

## Suppressing the .onAttach() local-VLM hint

The `.onAttach()` hook prints a one-line hint suggesting
`setup_local_vlm("light")` whenever Ollama is detected but
`gemma4:e2b` is not yet pulled. Suppress with:

```{r}
options(soilKey.suggest_local_vlm = FALSE)
```

Or, opt into automatic background pull:

```{r}
options(soilKey.auto_setup_vlm = TRUE)
# .. or via env var
Sys.setenv(SOILKEY_AUTO_SETUP_VLM = "1")
```

When the auto-setup option is enabled, soilKey runs `ollama pull
gemma4:e2b` in the background on attach. **CRAN-compliance:** the
hook never modifies the system without an explicit user opt-in
(option / env var).
